Towards a wide-coverage grammar for Swedish

### 2011-12-16 doc/report/report.tex
report on grammar and ramonas comments

\begin{document}
% ska gå att förstå och locka till läsning
%ska tydligt beskriva vad rapporten handlar om
%ska innehålla vettiga sökord
\title{Scaling up a GF grammar for free parsing} %A nice titleScaling up a GF grammar and preparing it for cool stuff for Swedish}
\author{Malin Ahlberg}
\maketitle
\newpage

@@

\tableofcontents
\newpage
\chapter{Introduction}
% more intro to gf? Good starting sentence.
The grammar formalism GF has so far been successful used for 
controlled languages, but recent experiments have showed
that it is also possible to use the framework for parsing free language.
Parsing, that is the task of automatically identifying morphological and
syntactical structures to text, is subject to increasing interest, especially
considering the steadily growing amount of data available online. 
So far there is no freely available grammar-driven parser that gives a deep analyse for
Swedish. The fact that %the existing rule-based Swedish
a parser or grammar is not freely available does not only restrict its
usage but also its possibilities of being developed. Our
goal is
to create a open source Swedish grammar and parser that can be continuously enhanced and
made use of in other projects.\\
To build a parser or a grammar from scratch would be not only time consuming but
also a great theoretical and technical challenge. In order to not reinvent the wheel
we develop our parser by combining well-tested resources.
We start from GF resource grammar and thereby get
%The usage of GF allows us to start from
a well-defined system for describing grammar, as well as a strong connection to
more than 20 other languages implemented in the framework. Further, we use
the extensive lexicon Saldo and the treebank Talbanken.
%>>The grammar defines a parser, which accepts all constructions defined by the grammar.

@@

%> dependent and implicit types : agda. (read gf book and krasses book)
A grammar written in GF can be used for both parsing and generation.
The parsing algorithm is incremental and has polynomial time and space
complexity \cite{gfMech}. 
The GF package also provides various tools for working with and using the grammars:
a compiler, an interpreter and a runtime system.
The grammars can be compiled into portable grammar format (PGF) \cite{pgf},
supported by %There are libraries for using PGF files with
Java, Java script, C and Haskell libraries. 
%>> Haskell only complete, also or C\# and C.  %may also be used in Android applications.
The interpreter, the GF shell, allows
the user to test grammars by commands for parsing, visualization of parse trees,
random generation, word alignment, morphological quizzes etc.
%The following example shows how to parse a sentence and visualizing the parse tree.
%, by using the \textit{pipe} \verb-|-, sending the result to 
\begin{figure}[h]
\begin{verbatim} 
> parse "jag ser katten" | visualize_parse -format=pdf -view=evince
> parse "jag ser katten" | visualize_tree  -format=pdf -view=evince
\end{verbatim}
\caption{Example of how the GF shell can be used. The first command will parse the given
sentence and visualize the resulting parse tree and the second will visualize the 
abstract syntax tree.}
\label{fig:shellvp}
\end{figure}
\begin{figure}[h]
\centering
\subfloat[GF abstract tree]{\label{pic:gfAtree}\includegraphics[width=100mm]{gfTree.pdf}}
\subfloat[Swedish parse tree]{\label{pic:gfStree}\includegraphics[width=30mm]{gfSTree.png}}
\caption{Parse tree and abstract tree for the sentence ``Jag ser katten".}
\label{fig:gftree1}
\end{figure}

@@

%>>the scope is controlled language, but possible ambiguities (discuss this in future
%>>work, saldo, conclusion). Ambiguities: from multilinguality, or inherited:
%>>syntactical (man with the telescope), lexical (youSg/Pl, växa).
%>> have ambiguity section. 
\newpage %as long as english picture bråks.
The scope of GF grammars is usually controlled language. By
restricting the coverage, the number of ambiguities
can be limited and we can ensure that the semantics is preserved during
translation. Inherited ambiguities and the ones arising from multilinguality will
of course remain, but can be controlled more easily. %Example of domainspecific word?
The use of controlled language thus gives a possibility of good and reliable translation
and GF has successfully been used in project like %Some examples of projects using GF are 
WebAlt \cite{webalt}, a project which aims to develop language independent
material for mathematics problems, and the formal software verification tool
KeY \cite{key}, in which GF is used for translating specification to natural language.
%>> key, talk, hats, bom (see references in Ramonas thesis).
%>> use talk here and move key to other place??
The framework is also used in the
European project MOLTO \cite{Molto} for online translation between up to 15 language.\\

\subsection{Writing a GF grammar}
The key feature in GF is the distinction between
\textit{abstract} and \textit{concrete} syntax. The abstract syntax represents
the internal structure and models the semantics without concern for language
specific features such as agreement or word order.
An abstract grammar can be implemented by a set of concrete grammars, each
representing one language. As a comparison, the abstract and concrete syntax
may be thought of as fstructures and cstructures in Lexical Functional Grammar (Bresnan and Kaplan, 1982).

@@

\subsection{The resource library}
\label{sec:resources}
%>>Best part? Biggest library?
%>>Most general resource, first large grammar, covers fundamental syntactics.
The GF package provides a advantageous resource library \cite{gf-resource}, covering the
fundamental morphology and syntactics of more than 20 languages.
%is a very useful part of the framework. It is provided in the 
%, they give a good start for all who want to write an application grammar.
%of a language for each application, all
%this can be found as libraries. So far there are 20 languages in the library, and
The languages all share the same abstract and are valuable
%they constitute  a very useful resource 
when implementing application grammars. There is also a small test lexicon
included, containing a few hundred common words.

@@

\subsection{Frontiers of Grammatical Framework}
As an open source-project, GF is constantly being developed and improved. New
languages are added, the compiler is being improved, ways of using it in more 
efficient and easygoing manners are added
and the possibilities to use GF in different environments
increased. There are research on how to make more use of the dependent
types, for reasoning by using ontologies \cite{ramona} or generating natural
language via Montague semantics \cite{montague}.
Further, experiments on using GF for free parsing has been conducted
\cite[\textsection 4.1]{gfMech}.
This is also a future goal for our project and one of the subgoal has been to
model an important part of Swedish in GF. The starting point was the resource
grammar to which the result has a strong connection.

@@

In the TAG formalism (Joshi,1975), there are projects on getting open source, wide-coverage grammars
for ie. English and Korean, but, to our knowledge, not for Swedish.  
%However has been two, about CLE \cite{cle} and what happened to it. Logic and translation
%to Swedish, S-CLE (Gambäck,1997) full analyse and semantics, translation,
%but coverage is too limited (nivres papper), The
%spoken lang. translator \cite{cle2}.

@@

\section{Results}
All words from Saldo were not be imported. There are a number of reasons
for this, the most obvious one is that we do not want all categories from
Saldo in the GF lexicon. Prepositions(\emph{pp}), numerals(\emph{nl}),
personal pronouns etc.
are assumed to be present in the resource grammars and should not be added again.
Saldo contains may other pronouns,
but they are not analysed the same way in GF (see section \ref{sec:swegf}) and
can thus not simply be extracted -- we first need to find their correct GF category.
Some experiments on finding
the right GF-category for those have been done using Talbanken, see p. \pageref{sec:gf.quant}.

@@

\chapter{The grammar}
\label{sec:prog.grammar}
An important part of this project has been to developed a grammar and
adapting it to cover
constructions used in Talbanken. A grammar can never be expected to give
full coverage of a language, but we aim for a grammar fragment which gives a deep
analysis of the most important Swedish constructions.
The starting point has been the GF resource grammar and our new grammar is
still compatible with this.
Before describing the work progress we will give an introduction to the
resource grammars in general and to the Swedish implementation in particular.
This is found in section \ref{sec:swegf} while section \ref{sec:Added} shows and
explains the development during this project.

@@

\section{The Swedish resource grammar}
%>> start by easy examples, jag ser inte dig, du ser inte mig osv. 
%>> plocka från svenska delen 1 2 4 
%>> show simple trees
%>> explain all in Sw. section.
%>> show abstract and concrete tree, compare. (or in Gf sect)
%>> no subjects or object, analyse is more from the lexicon,
%>> compare to others, ask elisabet.\\
\label{sec:swegf}
The GF resource grammars contains a fundamental description of Swedish,
covering the morphology and many phrase and clause constructions. Most of the implementation
is shared with the other Scandinavian languages.
Apart form describing morphology, the resource grammar deals with
word order, agreement, tense, basic conjunction etc.

All categories need tables covering the forms the phrase can be used in. They also
need to carry information about inherent features that may matter to other phrases.
Section \ref{gf:swe} contained an easy example with nouns. We used the categories \verb-N-
and \verb-N'- to separated between nouns with repsectively without a set number.
The full analysis includes a separation between the categories \verb-N- -- simple nouns --
and 
\verb-CN-, common noun phrases which may be modified by adjectival phrases or conjucted.
A \verb-CN- may be turned into a \verb-NP- by the rule \verb-DetNP- which sets
the number and definitness. Since
all morphological information is generated by the smart paradigms and contained in the \verb-N-
the construction of a \verb-NP- more or less consists of putting the pieces together.
The two basic articles are \verb-IndefArt- and \verb-DefArt- which
uses the post-modified form of the noun: katten.
However, we need to consider the rules described in section \ref{sec:swedishnoun}. If we have
a common noun phrase consisting of the parts \emph{`liten'} (\emph{`small'}) and \emph{`katt'}
(\emph{`cat'}), there are three ways they can be combined:
\begin{verbatim}
DetNP : Det -> CN -> NP ;
        en    liten katt     -- DIndef
        den   lilla katten   -- DDef Def
        min   lilla katt     -- DDef Indef
\end{verbatim}        
Which combination that shoud be used is decided by the determiner. Hence all determiners
in our grammar must keep information of which definitness they allow. They record field
has a field storing this and the parameters are called \verb-DIndef-, 
\verb-DDef Indef- and \verb-DDef Def-. The determiners have arguments telling
which number the noun phrase should be in.
We also need to make sure that we do not allow constructions as given in
\ref{sent:trottkatt}a.
The definite article  \emph{``en"}is considered to be a quantifier, in plural it
is the either \emph{``de"} or nothing:
\begin{verbatim}
DetNP : Det              -> CN        -> NP ;
        definite, plural +  katt      =  katter
        definite, plural +  stor katt =  de stora katterna
\end{verbatim}        
This difference depends on the common noun phrase, which must carry information
about whether it has been modified or not.
The resource grammar have differenciate between quantifiers -- \verb-Quant-,
determiners \verb-Det- and predeterminers. Predeterminers modify \verb-NP-s
while the other to modifies \verb-CN-s. The differences are shown in table
\ref{tab:detquant}
\begin{figure}[h]
\begin{tabular}{l|lll}
                        &\textbf{Has number}& \textbf{Has definitness}& Example  \\ 
\hline
\textbf{Predeterminers} & --  & --  & alla katter, all maten \\ 
\textbf{Quantifiers}&-- &\checkmark  & min katt, mina katter, *min katten \\ 
\textbf{Determiners}   &\checkmark &\checkmark & varje katt, *varje katten  \\ 
\end{tabular}
\caption{}\label{tab:detquant}
\end{figure}

@@

%Categories as V,V2,VS...
%Maybe talk about ställa.
Swedish verbs may take up to five arguments \cite[p. xx]{stymne}.
\enumsentence{Jag tar med mig den till honom}\label{sent:tamed}
These may be particles, reflexive object, indirect objects and direct objects.
In the grammmar these will be encoded in different ways. The particles are
given in the lexicon, as well as the prepositions that are choosen by the verbs.
The entry for sentence \ref{sent:tamed} looks like this, assuming that we already
have an constant \verb-take_V- which produces the correct conjugation for ``ta":
\verb|ta\_med\_V3 = dirV3 (reflV (partV (take\_V "med")) (mkPrep "till") ;|
The category \verb-V3- tells us that the verb takes two arguments (``den" and ``honom").
The preposition is set to ``till", reflV shows that the verb always is used with a reflexive
pronoun and partV gives the particle.
The fact that the choosen prepositions is given to the verb in the lexicon makes the 
parse tree visualisation algorithm put them together. See parse tree \ref{fig:tittapa}.
This is also the case for particles, see parse tree \ref{tittapart}, but even if
the visualizations look the same the two cases are treated differently inside the grammar
to keep the actual difference.
\enumsentence{På pojken tittar du \\ *På när du springer tittar jag}
As already stated, the visualized parse trees is not a complete represtation, and the reader
should not be confused or angry at this.\\

@@

\enumsentence{\shortexnt{6}{Han & ser &katten & han  &tycker & om}
              {he & sees & the cat & he & likes &}}
              %{`She sees the sleeping boy'}}
In addition to the basic resource grammar, additional constructions had been added
to the module \verb|Extra|. The functions given here do not have to be translatable
to all other language, but are meant to cover language specific constructions.
%For example, the \verb|Extra| module included
Among those were functions for topicalisation, although not all
of them were implemented for Swedish.
%word order, agreement, tempus, adverbs, adjectivs...
%and relatives
%Also info in the Extra module, where more language specific constructions
%can be found.
For example, \verb|FocObj| fronted the object as in sentence \ref{gfSwe:apple}.
\enumsentence{\shortex{6}
{Det & äpplet & vill & jag & inte & ha}
{that & apple & want & I & not &}
{`I don't want that apple'}}\label{gfSwe:apple}
The module provided functions for preposition stranding:
\enumsentence{\begin{tabular}{ll}
Stranded preposition & cf.\\
Vem måste jag akta mig för? & För vem måste jag akta mig?\\
Who do I need to watch out for? & For whom do I need to watch out?\\
\end{tabular}}

@@

\section{Development of the grammar}
\label{sec:Added}
The resource grammar is meant to cover the most important grammatical structures
present in all? languages. We have increased this by adding language specific
constructions. Talbanken has been our inspiration.
From the things listed in section \ref{sec:swedish}, the two first are basic
enough to be covered by the resources. The others have been added.

%%PASSIV
\subsection{The s-passive}
Passive voice is often used in Swedish, especially the
\textit{s-passive}.
\enumsentence{
\shortex{5}{Uppsatsen & skrev\textbf{s} & av & en & student.}
        {the essay & wrote+\textbf{s} & by & a & student}
        {`The essay was written by a student.'}}\label{sent:skrevs}
%{b. &En & student & skrev & uppsatsen.}
%        {&a &student &wrote &the essay}
%        {~~`A student wrote the essay.'}}
Some studies suggest that the s-passive is used in more than 80\% of the times \cite{laanemets}.
It is not as common in the other Scandinavian languages however, 
and it can not be used in all tenses for all words. The Norwegian 
translation of sentence \ref{sent:skrevs} is:
%The periphrastic passive is preferred, especially in spoken langauge.
\enumsentence{\begin{tabular}{ll}
Oppgaven ble skrevet av en student & [NO]\\
Uppsatsen blev skriven av en student& [SE]\\
\end{tabular}}
The corresponding Swedish sentence is acceptable, but not as natural sounding as sentence
\ref{sent:skrevs}.
The resource grammar for Scandinavian therefore ~~expressed~~ the passive using auxiliary verb. \\
\verb|PassV2 : V2  -> VP | \\
\verb|         ta -> blev tagen| \\
%>> how are they used with reflexives??
This rules allows two-place verbs to be used in passive by using \emph{bli} (\emph{become}), and thereby
turned into complete verb phrases; they no longer need an object.
The s-passive has now been added as the standard case for the Swedish GF
grammar. Periphrastic passive is still allowed, but as an alternative rather
than the default.
\verb|PassVPSlash : VPSlash  -> VP | \\
\verb|              ta -> togs| \\
\verb|              erbjöd -> erbjöds| \\
The grammar also allows not only V2 but all verb phrases missing an object to 
form passives:
\enumsentence{First place in two-place verbs\\\shortex{3}
{Jobbet & erbjöds & henne }
{the job & offered+\textbf{s}& henne}
{``The job was offered to her"}}
\label{ex:passV32}
\enumsentence{Second place in two-place verbs\\\shortex{3}
{Hon & erbjöds & jobbet}
{she & offered+\textbf{s}& the job}
{``She was offered the job"}}
\label{ex:passV33}
\enumsentence{\shortex{3}
{Huset & målades & rött}
{the house & painted+\textbf{s}& red}
{``The house was painted red"}}
\label{ex:passV2A}

\subsection{Impersonal constructions}%Presentering, , Topicalisering
\label{sec:Formal}
Formal subjects \cite[19]{SAG} is often used in Swedish.
\enumsentence{\shortex{6}
{Det & sitter & en & fågel & på & taket}
{It & sits & a & bird & on & the roof}
{``There is a bird sitting on the roof''}}
\emph{`Det'} has the position of the subject, and the real subject, 
\emph{`en fågel'} the one of an object.

@@

%'det är kul att..' Näst vanligaste två-ordskombinationen i talad svenska.
%%
%embedded - relative, 'sådan att' not really nice for normal swedish, and no agreement
%between subject 'en katt sådan att det regnar'. Kan säga 'en sån katt som sover, med 'sådan' + Rel
%har lagt till Det äpple + Rel (RelCNNP). *Det äpple är rött. men Det äpple jag äter är rött.
% 
\subsection{Formalizing the rules for reflexive pronouns by using dependent types}
%varandra and sin/sitt/sig.
An important piece in a Swedish grammar is the reflexive pronouns and
%both \emph{sin} and , eg. \emph{sin}.
the possessive pronouns as described on p. \pageref{swe:refl}.
The reflexive pronouns require an antecedent with which they agree in
number and person.
Our grammar should accept the following sentences:

>>>>

\begin{document}
% ska gå att förstå och locka till läsning
%ska tydligt beskriva vad rapporten handlar om
%ska innehålla vettiga sökord
\title{Towards a wide-coverage grammar for Swedish} %A nice titleScaling up a GF grammar and preparing it for cool stuff for Swedish}
\author{Malin Ahlberg}
\maketitle
\newpage

@@

\tableofcontents
\newpage
\chapter{Introduction}
% more intro to gf. dependently typed...high-level formalism
The grammar formalism GF \cite{gfbok} has so far been successfully used for 
controlled languages \cite{cnl}, but recent experiments have showed
that it is also possible to use the framework for parsing free language \cite{patent}.
 >> although the success rate is going steeply down when using uncontrolled nl. Better with annotated.
 --> sentence about connection GF - grammar based parser.
Parsing, that is the task of automatically identifying morphological and
syntactical structures to text, is subject to increasing interest, especially
considering the steadily growing amount of data available online. 
So far there is no freely available grammar-driven parser that gives a deep analyse for
Swedish. The fact that %the existing rule-based Swedish
a parser or grammar is not freely available does not only restrict its
usage but also its possibilities of being developed. Our
goal is
to create a open source Swedish grammar and parser that can be continuously enhanced and
made use of in other projects.\\
To build a parser or a grammar from scratch would be not only time consuming but
also a ~~great theoretical and technical challenge~~.
>> not duplicate, already existing resources.
In order to not reinvent the wheel
we develop our parser by combining well-tested resources.
We start from GF resource grammar and thereby get
%The usage of GF allows us to start from
a well-defined system for describing grammar, as well as a strong connection to
more than 20 other languages implemented in the framework. Further, we use
the extensive lexicon Saldo and the treebank Talbanken.
%>>The grammar defines a parser, which accepts all constructions defined by the grammar.

@@

%> dependent and implicit types : agda. (read gf book and krasses book)
A grammar written in GF can be used for both parsing and generation.
The parsing algorithm is incremental and has polynomial time and space
complexity \cite{gfMech}. 
The GF package also provides various tools for working with and using the grammars:
a compiler, an interpreter and a runtime system.
The grammars can be compiled into portable grammar format (PGF) \cite{pgf},
supported by %There are libraries for using PGF files with
Java, Java script, C and Haskell libraries. 
%>> Haskell only complete, also or C\# and C.  %may also be used in Android applications.
The interpreter, the GF shell, allows
the user to test grammars by commands for parsing, visualization of parse trees,
random generation, word alignment, morphological quizzes etc.
>> shell reference, servers, cloud. \footnote{http://www.grammaticalframework.org/demos/gfse/}
\begin{figure}[h]
\begin{verbatim} 
> parse "jag ser katten" | visualize_parse -format=pdf -view=evince
> parse "jag ser katten" | visualize_tree  -format=pdf -view=evince
\end{verbatim}
\caption{Example of how the GF shell can be used. The first command will parse the given
sentence and visualize the resulting parse tree and the second will visualize the 
abstract syntax tree.}
\label{fig:shellvp}
\end{figure}
\begin{figure}[h]
\centering
\subfloat[GF abstract tree]{\label{pic:gfAtree}\includegraphics[width=100mm]{gfTree.pdf}}
\subfloat[Swedish parse tree]{\label{pic:gfStree}\includegraphics[width=30mm]{gfSTree.png}}
\caption{Parse tree and abstract tree for the sentence ``Jag ser katten".}
\label{fig:gftree1}
\end{figure}

@@

%>>the scope is controlled language, but possible ambiguities (discuss this in future
%>>work, saldo, conclusion). Ambiguities: from multilinguality, or inherited:
%>>syntactical (man with the telescope), lexical (youSg/Pl, växa).
%>> have ambiguity section. 
\newpage %as long as english picture bråks.
The scope of GF grammars is usually controlled language. By
restricting the coverage, the number of ambiguities
can be limited and we can ensure that the semantics is preserved during
translation. Inherited ambiguities and the ones arising from multilinguality will
of course remain, but can be controlled more easily. %Example of domainspecific word?
The use of controlled language thus gives a possibility of good and reliable translation
and GF has successfully been used in project like %Some examples of projects using GF are 
WebAlt \cite{webalt}, a project which aims to develop language independent
material for mathematics problems, and the formal software verification tool
KeY \cite{key}, in which GF is used for translating specification to natural language.
%>> key, talk, hats, bom (see references in Ramonas thesis).
%>> use talk here and move key to other place??
>> talk!
The framework is also used in the
European project MOLTO \footnote{http://www.molto-project.eu/} for online
translation between up to 15 language.\\

\subsection{Writing a GF grammar}
The key feature in GF is the distinction between
\textit{abstract} and \textit{concrete} syntax. The abstract syntax represents
the internal structure and models the semantics without concern for language
specific features such as agreement or word order.
An abstract grammar can be implemented by a set of concrete grammars, each
representing one language. As a comparison, the abstract and concrete syntax
may be thought of as f-structures and c-structures in Lexical Functional Grammar (Bresnan and Kaplan, 1982).

@@

\subsection{The resource library}
\label{sec:resources}
%>>Best part? Biggest library?
%>>Most general resource, first large grammar, covers fundamental syntactics.
The GF package provides a advantageous resource library \cite{gf-resource}, covering the
fundamental morphology and syntax of more than 20 languages.
%is a very useful part of the framework. It is provided in the 
%, they give a good start for all who want to write an application grammar.
%of a language for each application, all
%this can be found as libraries. So far there are 20 languages in the library, and
The languages all share the same abstract and are valuable
%they constitute  a very useful resource 
when implementing application grammars. There is also a small test lexicon
included, containing a few hundred common words.

@@

\subsection{Frontiers of Grammatical Framework}
As an open source-project, GF is constantly being developed and improved. New
languages are added, the compiler is being improved, ways of using it in more 
efficient and easygoing manners are added
and the possibilities to use GF in different environments
increased. There are research on how to make more use of the dependent
types, for reasoning by using ontologies \cite{ramona}and krasse?? or generating natural
language via Montague semantics \cite{montague}.
Further, experiments on using GF for free parsing has been conducted
\cite[\textsection 4.1]{gfMech}.
This is also a future goal for our project and one of the subgoal has been to
model an important part of Swedish in GF. The starting point was the resource
grammar to which the result has a strong connection.

@@

In the TAG formalism (Joshi,1975), there are projects on getting open source, wide-coverage grammars
for ie. English and Korean, but, to our knowledge, not for Swedish.

ParGram! More parallel development than Lingo.

%However has been two, about CLE \cite{cle} and what happened to it. Logic and translation
%to Swedish, S-CLE (Gambäck,1997) full analyse and semantics, translation,
%but coverage is too limited (nivres papper), The
%spoken lang. translator \cite{cle2}.

@@

\section{Results}
\label{sec:saldoRes}
All words from Saldo were not be imported. There are a number of reasons
for this, the most obvious one is that we do not want all categories from
Saldo in the GF lexicon. Prepositions(\emph{pp}), numerals(\emph{nl}),
personal pronouns etc.
are assumed to be present in the resource grammars and should not be added again.
Saldo contains may other pronouns,
but they are not analysed the same way in GF (see section \ref{sec:swegf}) and
can thus not simply be extracted -- we first need to find their correct GF category.
Some experiments on finding
the right GF-category for those have been done using Talbanken, see p. \pageref{sec:gf.quant}.

@@

\chapter{The grammar}
\label{sec:prog.grammar}
An important part of this project has been to developed a grammar and
adapting it to cover
constructions used in Talbanken. A grammar can never be expected to give
full coverage of a language, but we aim for a grammar fragment which gives a deep
analysis of the most important Swedish constructions.
The starting point has been the GF resource grammar and our new grammar is
still compatible with this.
Before describing the work progress we will give an introduction to the
resource grammars in general and to the Swedish implementation in particular.
This is found in section \ref{sec:swegf} while section \ref{sec:Added} shows and
explains the development during this project.
>> the grammar will not be totally covering, but interesting and useful.

@@

\section{The Swedish resource grammar}
%>> start by easy examples, jag ser inte dig, du ser inte mig osv. 
%>> plocka från svenska delen 1 2 4 
%>> show simple trees
%>> explain all in Sw. section.
%>> show abstract and concrete tree, compare. (or in Gf sect)
%>> no subjects or object, analyse is more from the lexicon,
%>> compare to others, ask elisabet.\\
\label{sec:swegf}
>> diagram of modules, etc.
>> gf has pos-driven, leads to part-of-sentence analyse. first have v2,a2,vs etc.
>> they may be combined in traditional way, and so the types depend and relate to each other
>> see slides from summer school.
>> functions for predication, complementation.
>> rather different from others syntactical theories. does not remember syntactical
>> parts of sentences, like subj, obj, attributes.
>> worked well until now, but particular langs. needs particular solutions. --> reflexives!
>> no syntactic theory can cover a whole language, and so gf can't either.
>> shares 80\% with scandinavian, mostly words that differ.
The GF resource grammars contains a fundamental description of Swedish,
covering the morphology and many phrase and clause constructions. Most of the implementation
is shared with the other Scandinavian languages.
Apart form describing morphology, the resource grammar deals with
word order, agreement, tense, basic conjunction etc.

All categories need tables covering the forms the phrase can be used in. They also
need to carry information about inherent features that may matter to other phrases.
Section \ref{gf:swe} contained an easy example with nouns. We used the categories \verb-N-
and \verb-N'- to separated between nouns with repsectively without a set number.
The full analysis includes a separation between the categories \verb-N- -- simple nouns --
and 
\verb-CN-, common noun phrases which may be modified by adjectival phrases or conjucted.
A \verb-CN- may be turned into a \verb-NP- by the rule \verb-DetNP- which sets
the number and definitness. Since
all morphological information is generated by the smart paradigms and contained in the \verb-N-
the construction of a \verb-NP- more or less consists of putting the pieces together.
The two basic articles are \verb-IndefArt- and \verb-DefArt- which
uses the post-modified form of the noun: katten.
However, we need to consider the rules described in section \ref{sec:swedishnoun}. If we have
a common noun phrase consisting of the parts \emph{`liten'} (\emph{`small'}) and \emph{`katt'}
(\emph{`cat'}), there are three ways they can be combined:
\begin{verbatim}
DetNP : Det -> CN -> NP ;
        en    liten katt     -- DIndef
        den   lilla katten   -- DDef Def
        min   lilla katt     -- DDef Indef
\end{verbatim}        
>> compare NP to HPSG.
>> CN may be different in other theories.
>> how we have basic Pred, compl
>> clase is a pre-sentence needs tense, polarity etc.
Which combination that shoud be used is decided by the determiner. Hence all determiners
in our grammar must keep information of which definitness they allow. They record field
has a field storing this and the parameters are called \verb-DIndef-, 
\verb-DDef Indef- and \verb-DDef Def-. The determiners have arguments telling
which number the noun phrase should be in.
We also need to make sure that we do not allow constructions as given in
\ref{sent:trottkatt}a.
The definite article  \emph{``en"}is considered to be a quantifier, in plural it
is the either \emph{``de"} or nothing:
\begin{verbatim}
DetNP : Det              -> CN        -> NP ;
        definite, plural +  katt      =  katter
        definite, plural +  stor katt =  de stora katterna
\end{verbatim}        
This difference depends on the common noun phrase, which must carry information
about whether it has been modified or not.
The resource grammar have differenciate between quantifiers -- \verb-Quant-,
determiners \verb-Det- and predeterminers. Predeterminers modify \verb-NP-s
while the other to modifies \verb-CN-s. The differences are shown in table
\ref{tab:detquant}
\begin{figure}[h]
\begin{tabular}{l|lll}
                        &\textbf{Has number}& \textbf{Has definitness}& Example  \\ 
\hline
\textbf{Predeterminers} & --  & --  & alla katter, all maten \\ 
\textbf{Quantifiers}&-- &\checkmark  & min katt, mina katter, *min katten \\ 
\textbf{Determiners}   &\checkmark &\checkmark & varje katt, *varje katten  \\ 
\end{tabular}
\caption{}\label{tab:detquant}
\end{figure}

@@

%Categories as V,V2,VS...
%Maybe talk about ställa.
Swedish verbs may take up to four arguments \cite[p. 53]{stymne}.
\enumsentence{Jag tar med mig den till honom}\label{sent:tamed}
These may be particles, reflexive object, indirect objects and direct objects.
In the grammmar these will be encoded in different ways. The particles are
given in the lexicon, as well as the prepositions that are choosen by the verbs.
The entry for sentence \ref{sent:tamed} looks like this, assuming that we already
have an constant \verb-take_V- which produces the correct conjugation for ``ta":
\verb|ta\_med\_V3 = dirV3 (reflV (partV (take\_V "med")) (mkPrep "till") ;|
The category \verb-V3- tells us that the verb takes two arguments (``den" and ``honom").
The preposition is set to ``till", reflV shows that the verb always is used with a reflexive
pronoun and partV gives the particle.
The fact that the choosen prepositions is given to the verb in the lexicon makes the 
parse tree visualisation algorithm put them together. See parse tree \ref{fig:tittapa}.
This is also the case for particles, see parse tree \ref{tittapart}, but even if
the visualizations look the same the two cases are treated differently inside the grammar
to keep the actual difference.
\enumsentence{På pojken tittar du \\ **På när du springer tittar jag}
As already stated, the visualized parse trees is not a complete represtation, and the reader
should not be confused or angry at this.\\

@@

\enumsentence{\shortexnt{6}{Han & ser &katten & han  &tycker & om}
              {he & sees & the cat & he & likes &}}
              %{`She sees the sleeping boy'}}
In addition to the basic resource grammar, additional constructions had been added
>> core is shared but particular lang. has particular features.
to the module \verb|Extra|. The functions given here do not have to be translatable
to all other language, but are meant to cover language specific constructions.
Among those were functions for topicalisation, although not all
of them were implemented for Swedish.
%word order, agreement, tempus, adverbs, adjectivs...
%and relatives
%Also info in the Extra module, where more language specific constructions
%can be found.
For example, \verb|FocObj| fronted the object as in sentence \ref{gfSwe:apple}.
\enumsentence{\shortex{6}
{Det & äpplet & vill & jag & inte & ha}
{that & apple & want & I & not &}
{`I don't want that apple'}}\label{gfSwe:apple}
The module provided functions for preposition stranding:
\enumsentence{\begin{tabular}{ll}
Stranded preposition & cf.\\
Vem måste jag akta mig för? & För vem måste jag akta mig?\\
Who do I need to watch out for? & For whom do I need to watch out?\\
\end{tabular}}

@@

\section{Development of the grammar}
\label{sec:Added}
The resource grammar is meant to cover the most important grammatical structures
present in all? languages. We have increased this by adding language specific
constructions. Talbanken has been our inspiration.
From the things listed in section \ref{sec:swedish}, the two first are basic
enough to be covered by the resources. The others have been added.
>> hard to estimate before, since no big use case/resource.
>> after initial experiments, first evaluation. found much room for improvement. 
>> complete morphology, limited syntax
>> discovered constructions missing.

%%PASSIV
\subsection{The s-passive}
Passive voice is often used in Swedish, especially the
\textit{s-passive}.
\enumsentence{
\shortex{5}{Uppsatsen & skrev\textbf{s} & av & en & student.}
        {the essay & wrote+\textbf{s} & by & a & student}
        {`The essay was written by a student.'}}\label{sent:skrevs}
%{b. &En & student & skrev & uppsatsen.}
%        {&a &student &wrote &the essay}
%        {~~`A student wrote the essay.'}}
Some studies suggest that the s-passive is used in more than 80\% of the times \cite{laanemets}.
It is not as common in the other Scandinavian languages however, 
and it can not be used in all tenses for all words. The Norwegian 
translation of sentence \ref{sent:skrevs} is:
%The periphrastic passive is preferred, especially in spoken langauge.
\enumsentence{\shortexnt{7} %\begin{tabular}{ll}
{Oppgaven &ble& skrevet& av& en& student & [NO]}
{uppsatsen& blev& skriven& av& en & student& [SE]}}
%\end{tabular}}
The corresponding Swedish sentence is acceptable, but not as natural sounding as sentence
\ref{sent:skrevs}.
The resource grammar for Scandinavian therefore ~~expressed~~ the passive using auxiliary verb. \\
\verb|PassV2 : V2  -> VP | \\
\verb|         ta -> blev tagen| \\
%>> how are they used with reflexives??
This rules allows two-place verbs to be used in passive by using \emph{bli} (\emph{become}), and thereby
turned into complete verb phrases; they no longer need an object.
The s-passive has now been added as the standard case for the Swedish GF
grammar. Periphrastic passive is still allowed, but as an alternative rather
than the default.
\verb|PassVPSlash : VPSlash  -> VP | \\
\verb|              ta -> togs| \\
\verb|              erbjöd -> erbjöds| \\
The grammar also allows not only V2 but all verb phrases missing an object to 
form passives:
\enumsentence{First place in two-place verbs\\\shortex{3}
{Jobbet & erbjöds & henne }
{the job & offered+\textbf{s}& henne}
{``The job was offered to her"}}
\label{ex:passV32}
\enumsentence{Second place in two-place verbs\\\shortex{3}
{Hon & erbjöds & jobbet}
{she & offered+\textbf{s}& the job}
{``She was offered the job"}}
\label{ex:passV33}
\enumsentence{\shortex{3}
{Huset & målades & rött}
{the house & painted+\textbf{s}& red}
{``The house was painted red"}}
\label{ex:passV2A}

\subsection{Impersonal constructions}%Presentering, , Topicalisering
\label{sec:Formal}
Formal subjects \cite[\textsection 19]{SAG} is often used in Swedish.
\enumsentence{\shortex{6}
{Det & sitter & en & fågel & på & taket}
{It & sits & a & bird & on & the roof}
{``There is a bird sitting on the roof''}}
\emph{`Det'} has the position of the subject, and the real subject, 
\emph{`en fågel'} the one of an object.

@@

%'det är kul att..' Näst vanligaste två-ordskombinationen i talad svenska.
%%
%embedded - relative, 'sådan att' not really nice for normal swedish, and no agreement
%between subject 'en katt sådan att det regnar'. Kan säga 'en sån katt som sover, med 'sådan' + Rel
%har lagt till Det äpple + Rel (RelCNNP). *Det äpple är rött. men Det äpple jag äter är rött.
% 
\subsection{Formalizing the rules for reflexive pronouns by using dependent types}
>> solution requires new strategies, combining 2 theories! (part of speech, part of sentence)
An important piece in a Swedish grammar is the reflexive pronouns and
the possessive pronouns as described on p. \pageref{swe:refl}.
The reflexive pronouns require an antecedent with which they agree in
number and person.
Our grammar should accept the following sentences:
